Foraging: The Triumphant Return

### 2022-07-07 manuscript.tex

Different aspects of foraging behaviour, such as the search image and the search path, clearly interact with one another, but most research has tried to understand them separately. In many cases, the pattern of foraging behaviour has also been studied using aggregate measures calculated on the level of a trial, such as the mean number of 'runs' (where the same type of target is selected multiple times in a row) or the total number of targets found during the longest run. However, we have recently developed a new model to analyse foraging data, based on a sampling without replacement procedure \cite{clarke2022}. By using this model, we demonstrated that we were able to break down foraging into a number of different cognitive biases, such as a preference to stick to the same target type, or a preference for nearby targets, and used this model to successfully re-analyse data from a number of open access datasets \cite{kristjansson2014, thornton2020, clarke2022, tagu2020}.

@@

Figure \ref{fig:cal2} hints at some individual differences when it comes to how predictable the model is: we can see a lot of variation in the size of the weights. However, interestingly, we find that this variability can be explained by our model parameters: nearly all of the differences between people, and between feature and conjunction conditions, can be explained by the $bP$ parameter, which is our proximity bias. The model is doing a better job with participants who show a stronger proximity bias.

@@

Figure \ref{fig:qjep_init_sel_hex} shows the distribution of the locations of first target selections. We can see that unlike the central bias in fixations during scene viewing, the distribution of initial target locations appear to be bimodal: while most targets are positioned in the top left hand corner, there is a second smaller distribution of central target selections. This appears to be due to different participants choosing to utilise different strategies rather than within-subject variation. However, we have no evidence that these different initial strategies affect the model parameters in any systematic way (see \textit{Supplementary Material}). We fit two dwill try two different modelling approaches (multi-level and mixture) to

@@

\begin{figure}[H]
\centering
\includegraphics[width=12 cm]{Figures/init_sel_mdl.pdf}
\caption{Posterior fits for (\textit{left}) horizontal and (\textit{middle}) vertical locations of the initial target selection. Each line represents a different participant. (\textit{right) A scatter plot showing the $x$ and $y$ posterior means for each participant.}
\label{fig:qjep_init_sel_mdl}
\end{figure}

@@

To compare the two approaches, we assess how well they fit the empirical data: we can use the fitted distributions to calculate the weight assigned to the selected target by each method for each trial in our dataset (Figure \ref{fig:qjep_init_sel_pred}). We can see that both models give a similar distribution of target weights and accuracies. Overall, both methods  select the correct target in around half of the trials. Interesting, as above, there is considerable variation between participants, with participants who favour the top-left corner being easier to predict.

@@

To what extent does our model predict behaviour on a trial-by-trial basis? Surprisingly well - this is a difficult task, in some cases the model can be over 80\% accurate. Are there any examples of similar models that can make such fine grained predictions?

What mistakes does the model make? 
Runner up item weights are a bit too low - this probably suggests a 'peakier' distribution for real human foragers, with only a few targets seriously considered, and everything else being approximately zero weighted?
Better at feature search compared to conjunction search, better at people with a stronger proximity bias - people who do unpredictable things are unpredictable? Does this suggest that some of our model parameters are less useful at predicting what occurs on a trial-by-trial basis (although they might be useful in aggregate).

To a large extent, initial item selection strategy seems to be stable - either you're someone who starts in the corner, or in the centre of the screen. This makes predicting where people start actually also quite accurate. Fancy multi-level model not really any better at predicting this than a mixture model. But this probably depends on the exact nature of the task (i.e. central bias not seen in \citep{kristjansson2014}, presumably because this was carried out on an iPad - affects people's behaviour in some way we don't really understand?)

Remaining open questions?
Incorporating other biases into initial target selection, e.g. $pA$?

>>>>

Different aspects of foraging behaviour, such as the search image and the search path, clearly interact with one another, but most research has tried to understand them separately. In many cases, the pattern of foraging behaviour has also been studied using aggregate measures calculated on the level of a trial, such as the mean number of 'runs' (where the same type of target is selected multiple times in a row) or the total number of targets found during the longest run. However, we have recently developed a new model to analyse foraging data, based on a sampling without replacement procedure \cite{clarke2022foraging}. By using this model, we demonstrated that we were able to break down foraging into a number of different cognitive biases, such as a preference to stick to the same target type, or a preference for nearby targets, and used this model to successfully re-analyse data from a number of open access datasets \cite{kristjansson2014, thornton2020, clarke2022, tagu_kristjansson_2021}.

@@

Figure \ref{fig:cal2} hints at some individual differences when it comes to how predictable the model is: we can see a lot of variation in the size of the weights. However, interestingly, we find that this variability can be explained by our model parameters: nearly all of the differences between people, and between feature and conjunction conditions, can be explained by the $bP$ parameter, which is our proximity bias. The model is doing a better job with participants who show a stronger proximity bias (see \textit{Supplementary Materials} for more details).

@@

Figure \ref{fig:qjep_init_sel_hex} shows the distribution of the locations of first target selections. We can see that unlike the central bias in fixations during scene viewing, the distribution of initial target locations appear to be bimodal: while most targets are positioned in the top left hand corner, there is a second smaller distribution of central target selections. This appears to be due to different participants choosing to utilise different strategies rather than within-subject variation. However, we have no evidence that these different initial strategies affect the model parameters in any systematic way (see \textit{Supplementary Material}). We try two different modelling approaches (multi-level and mixture) to assess whether the less complex mixture model is able to account for the patterns of starting positions seen.

@@

\begin{figure}[H]
\centering
\includegraphics[width=12 cm]{Figures/init_sel_mdl.pdf}
\caption{Posterior fits for (\textit{left}) horizontal and (\textit{middle}) vertical locations of the initial target selection. Each line represents a different participant. (\textit{right}) A scatter plot showing the $x$ and $y$ posterior means for each participant.}
\label{fig:qjep_init_sel_mdl}
\end{figure}

@@

To compare the two approaches, we assess how well they fit the empirical data: we can use the fitted distributions to calculate the weight assigned to the selected target by each method for each trial in our dataset (Figure \ref{fig:qjep_init_sel_pred}). We can see that both models give a similar distribution of target weights and accuracies. Overall, both methods  select the correct target in around half of the trials. Interestingly, as above, there is considerable variation between participants, with participants who favour the top-left corner being easier to predict.

@@

In our original model of visual foraging \citep{clarke2022foraging}, we were able to robustly measure aggregate parameters that underlie behaviour, at both an overall mean level across an experiment, and at the level of individual participants. In the current manuscript, we asked a more difficult question: could this model make target-by-target predictions, guessing which target would be picked next on any given trial? Despite this being a relatively difficult task, we found that our model was in some cases over 80\% accurate, demonstrating that at least in some cases, participants' behaviour in this task is remarkably predictable.

A benefit of our approach is that it is relatively easy to interrogate the model to try to understand why prediction accuracy is higher for some participants and trials than others. The model is generally better at feature search compared to conjunction search, and also seems to be better at predicting participants with a stronger proximity bias. The model also seems to assign more weight to 'runner-up' items than real humans do, perhaps suggesting a 'peakier' choice distribution for real foragers, with only a few targets seriously considered, and everything else being approximately zero weighted.

However, we can also delve deeper into exploratory hypotheses about other factors that might influence predictability. One suggestion is that the people who our model is able to predict are in some way the most 'optimal'. If this is the case, we might expect that model predictability should correlate with how good people are at the task, although we do not find any evidence that model accuracy correlates with reaction time on a by-participant or by-trial basis (see \textit{Supplementary Materials}). This may be due to the fact that this task was not time-limited, and instead participants were encouraged to be as accurate as possible (in fact, the trial restarted if the participant clicked on a non-target \citep{clarke2022}). However, there is some evidence that on a target-by-target basis, faster moves from one target to the next are more easily predicted by the model (see \textit{Supplementary Materials}). This likely is linked to the fact that our model is better at predicting participants with a stronger proximity bias, as targets that are closer to the previous target are also likely to be selected more quickly. Intuitively, it makes sense that the model struggles to make predictions when a participant has exhausted a local patch and may need to make a bigger jump to a more distant area to carry on collecting targets, as there may be multiple candidates and distance may no longer be such a good predictor (for example, people may choose to move to the next most densely populated patch).

There has been some previous work on within-trial behaviour in the context of foraging. For example, \citep{tagu2020} found that inter-target times vary across a trial (with targets selected later in the trial being slower) and inter-target distances also vary (with targets selected later in the trial having greater inter-target distances). Similarly, \citep{thornton2022foraging} showed that switches between target categories can be characterised by a change in foraging 'tempo', and if participants are forced to forage at a particular speed, by asking them to synchronise with an auditory signal, higher tempos led to a systematic decrease in the probability of participants switching between target categories. In the current version of the model, we do not consider time explicitly (although distance is likely to be highly correlated with time in this task). However, it would be possible to extend the model to incorporate timing information, and this may help to improve the target-by-target predictions.

From inspection of Figure \ref{fig:qjep_paths}, we noticed that in some cases, participants may be using a type of local path minimisation procedure, selecting a further away target in order to minimise overall path length, which the model does not always seem to predict. Could this help to explain some of the mistakes the model makes? Overall, the model doesn't seem to make substantially longer paths through the items than the human participants (see \textit{Supplementary Materials}). However, there is a relationship between model accuracy and the proportion of times that participants select paths shorter than the model: as model accuracy increases, people seem to be more likely to have paths shorter than the model predicts. This seems to suggest that path minimisation is not a good explanation for the targets our model is failing to capture, and instead the people that are difficult to predict are taking longer paths than the model thinks they should take.

Our exploratory analyses seem to hint at the idea that the people who are difficult for the model to predict are those who do not behave in a manner that optimises path length. It could be that these participants are simply unpredictable, and do not behave in a consistent manner on a target-by-target basis, making it difficult for a model to predict their behaviour at this level of granularity. However, it is possible that there are other factors that we have not taken into account that could further help explain these discrepancies. For example, it is possible that relatively unpredictable participants use multiple rules and switch between them in ways that we are not currently capturing in the model. Some participants may also lose track of where they are more frequently (perhaps due to attentional lapses) and thus may show foraging patterns that are predictable apart from unpredictable discontinuities, which again may not be easy to model. Finally, the displays had artificial discontinuities (i.e. screen edges) which may lead to different behaviour in different participants: some may carry out the task as if they are reading, reaching the right hand edge and then swinging back to the left, whereas others may 'bounce' off an edge they have just reached. Our model may well be able to predict the latter, but could struggle with the former behaviour.

To a large extent, initial item selection strategy seems to be stable within a participant, with people either starting in the corner or the centre of the screen. Our model is therefore also relatively accurate for selecting the initial target, selecting the correct target in around half of the trials (with a mixture model performing similarly to a multi-level model with many more parameters). Participants who prefer the top left corner were easiest to predict, probably because these participants formed the majority group in this experiment. Our results are congruent with \citep{kristjansson2022moving}, who found a bottom left bias in a 3D environment They suggest this may be an advantageous strategy because it allows your foraging to be more organised (e.g. in an 'S-shape', as has been found in real world search tasks \citep{riggs2017importance}). The smaller group of participants who prefer to start in the centre of the screen may be displaying similar behaviour to the well-known central bias in eye movements \citep{tatler2007central, tseng2009quantifying, clarke2014deriving} However, it is worth noting that these findings probably depend strongly on the context of the task. For example, there was no obvious central bias subgroup in \citep{kristjansson2014}, possibly reflecting the fact that this experiment was carried out on an iPad, compared to the computer screen and mouse used in \citep{clarke2022}. One challenge for future modelling work is to what extent we should try to account for these types of task-specific details: by incorporating them, we can generate better predictions for the experiment we are currently modelling, but perhaps at the cost of generalisability to other paradigms.

One of the key benefits of computational models is that they allow us to rigorously test how well we are able to predict behaviour, and can provide insights into what factors we are not capturing that may have important influences on how participants complete a task. We have demonstrated that our foraging model \cite{clarke2022foraging} is not only able to predict behaviour on aggregate, but can also make reasonable predictions at the target-by-target level. It is particularly good if a participant has a strong proximity bias, and on 'feature' trials where participants are normally switching between target types fairly frequently. It is also good at predicting where participants will start on a trial. The model finds it more difficult to predict 'discontinuous' jumps, which could be caused by a range of factors: a more difficult foraging task (e.g. conjunction searches), local path length minimisation, inattention, or aspects of the physical search space (e.g. edges of the screen). However, it would be possible to extend the model to incorporate these factors e.g. by incorporating a heuristic that would allow for local path minimisation, or introducing time as a predictor in the model. Our findings also suggest possible future directions for empirical work, such as evaluating the effect of inattention on foraging behaviour, or how the foraging targets are organised in space. We suggest that computational modelling is a powerful tool for helping us to understand behaviour, both by incorporating previous research into a shared framework, and by making testable predictions for future work.
